Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-04-26 paper/document.tex
Final draft of paper.

\begin{abstract}
What is a developer's contribution to a repository? By only counting commits and number of lines changed, existing tools that analyze source code repositories fall short on showing the effective contributions made by each developer. When many commits are viewed as a group, the details are lost. Commit information can be misleading since lines of code give no indication of what was actually being worked on without careful examination of the changed code. Providing a semantic view of this information could provide deeper insights into how software projects evolve since changes to design and features are not clearly visible from line changes alone. We present TypeV: a method for visualizing Java source code repositories. Instead of counting line changes in a commit we extract detailed type information over time by using the differences between abstract syntax trees (ASTs). We are then able to track the additions and deletions of declarations and invocations for each type. Furthermore, we can track each author's type usage over time. Using TypeV, we examine specific cases in well-known repositories where our tool reveals interesting and useful information. We then compare type coverage information from the AST compared to file coverage to determine if unique information is provided by type information.
\end{abstract}

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{antlr4-code-freq}
\caption{GitHub's code frequency graph for ANTLR4. Note that the graph was cropped to better show the details. Thus, the changes that occur in mid 2015 are much larger than shown. The additions reach over 400k and the deletions are over 300k.}
\label{fig:GH_code_freq}
\end{figure*}

@@

Source code is often represented as text in a source code file. This text is then parsed by a compiler or interpreter with the intent of eventually executing the code represented by the source text. While programmers may type in individual characters, their intention is at a much higher level. Their intent is written in tokens that represent statements, identifiers and types that are used to solve their problems. Programmers solve problems in a way that is readable to other developers and executable by a computer. The programmer rarely operates at the character level, or even the line level, so why should our analyses of source code operate at granularities unfamiliar to developers?
The plain text nature of software does not stop at source code level. The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which store the differences in lines of text in a file between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for Git repositories provide some basic tools for analyzing projects over time~\cite{github-graphs,bitbucket-graphs}; however, as we will show, these tools often lack the kind of detail to gather insightful analysis. As with source code, commits operate at the text level. While commits may store individual changes as lines of code, they imply a change in \emph{structure} of the software.  By visualizing lines changed, or files changed, the current state-of-the-art for software project visualization rarely capture the \emph{meaning} behind each commit; lines of code alone does not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed does not necessarily reflect meaningful changes to a repository~\cite{robles2014}.

@@

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualizes changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} Are type coverage file coverage different metrics?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

@@

Analyzing source code is a popular research topic and Java is often chosen as the subject due to its popularity. Grechanik \textit{et al.} \cite{Grechanik:2010:EIL:1852786.1852801} examined the structure of Java programs mined from 2080 programs. This paper examined the breakdown of syntactic structures in open source repositories, however it does not consider per author statistics. Parnin \textit{et al.} \cite{Parnin:2011:JGA:1985441.1985446} mined repositories to see how Java generics have been used in open source projects and found that generic usages were often introduced by a single developer in a project and were primarily being used for collecting and traversing lists of objects.

Further analysis of the Java language has been done using ASTs. L\"{a}mmel \textit{et al.} \cite{Lammel:2011:LAA:1982185.1982471} used ASTs to examine the usage of APIs in Java projects. They found what APIs are popular and if they were used in a framework like manner. Dyer \textit{et al.} \cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. The authors found the most popular features and how teams adopted new features. New features varied greatly in popularity and adoption rate. They did not check to see how much a developer used a feature but instead checked to see if they used it at all. In a more general analysis of programming languages, Meyerovich and Rabkin \cite{Meyerovich:2013:EAP:2509136.2509515} surveyed developers and examined repositories to learn about language adoption and usage. Developers feel that certain language features are more important than others.

The focus of these papers is on how programming languages and features are used by looking at many repositories. Java is a strongly typed language, yet there is no focus specifically on how types are used throughout a project. Type info may provide additional information on how developers use languages but has not been explored.

@@

A variety of visualizations tools get created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.} \cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directoryies and display the cardinality of a developer's commits Wu \textit{et al.} \cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies \cite{6980224} or files \cite{6650525, 7332421} over time.

Voinea \textit{et al.} \cite{voinea2006} visualizes CVS repositories using CVSgrab and adding their own enhancements to show changes to files over time. Files can be sorted by creation time, activity, or alphabetically and colour coded based on author, file size, file type or by search strings. Kim \textit{et al.} \cite{Kim:2009:DRS:1555001.1555046} built a tool called Logical Structural Diff (LSdiff) which examines code changes and produces logical rules to represent them. Changes which do not match the rules are then pointed out to the developer. They also formed a focus group with software engineers who found that LSdiff complements existing tools well and showed promise for discovering inconsistent changes.

@@

There is also research on how developers use software repositories to collaborate. Wagstrom \textit{et al.} \cite{Patrick:Wagstrom:2012} explored the roles that developers take in a networked, social development environments like GitHub. These role were defined as their level of contribution as well as the types of issues they dealt with. Developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects. Lee \textit{et al.} \cite{lee2013} developed a tool for visualizing the branching structure of git repositories. This was used to analyze concurrent workflows in popular open source repositories.

@@

Some visualization tools which are used for managing or making decisions about software rather than study the changes taking place. The service EasyBi \cite{EasyBi} provides additional visualizations for Git repositories for the purpose of business decisions. EasyBi works by taking the git log file and generating charts and graphs from it. It tries to solve the issue of summarizing many commits and has greater controls for granularity and authors but still operates on commits and line differences. The Gitinspector project \cite{Gitinspector} is an open source tool for the statistical analysis of git repositories. It provides an advanced timeline view of changes per author and cumulative statistics but again relies upon the line differences to summarize the work of each author.

The GitHub tools have the strength that they do not rely on what type of code is being committed. It does not matter what language or if errors exist in the commit, GitHub will simply spit out the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear breakdown of is being changed only that changes have occurred.

Consider tools easily available to an open source software developer; for example GitHub graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. Since GitHub tracks contributions based solely on number of commits and numbers of lines changes, data may be lacking, or even misleading. In the case of the ANTRL4 project (Figure~\ref{fig:GH_code_freq}), the graph creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to truncate the graph to be able to show details of other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by numerous reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps even a huge refactoring effort.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}).

An \emph{abstract syntax tree} (AST) is a data structure derived from source code that breaks its syntactic constructs into a tree. Each node in the tree represents a syntactically valid chunk of code. This captures the structure of the code in an abstract and tractable manner. Because of its ease of analysis, converting raw source text to an AST is often one of the first steps performed by a compiler, performed to facilitate semantic analysis of the source code. When a programmer makes a non-trivial change in code in a file, that difference will be reflected in the AST. Non-trivial changes are those that affect how a compiler will interpret the source code, and may have effects on the executable code. Trivial changes are those that have no affect on how a compiler will interpret the source text, such as changing insignificant whitespace or comments in source code. Thus, ASTs of consecutive revisions in a code repository can be compared to see what \emph{structures} a programmer is modifying.

@@

\begin{description}
\item [declaration] 
  Stating that use of a field or variable of a given type. For example declaration of a \texttt{String} type variable might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{String myString;} \\
  \end{adjustwidth}
\item [invocation] 
  A call site for a method that belongs to a type. For example an invocation of the trim method of a String type might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{myString.trim();} \\
  \end{adjustwidth}
\end{description}

@@

This states that two declarations of type \texttt{java.net.URI} were added, and two declarations of type \texttt{java.lang.String} were deleted. Note the fully qualified type names. The source text may not literally have the characters \texttt{java.net.URI}, but Spoon was able to infer from the import statements in the Java source text that references to ''URI'' resolve to the fully-qualified type name \texttt{java.net.URI}. Thus, TypeV is able to keep track of the actual class being used and is not confused when two classes in separate packages have the same name.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit we dumped the following information to a file: author's name, date of commit, commit id, commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited files we also compared the file at the time of the commit with its parent commit using the AST differencing as described above. We achieved this by using calling \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent. We were able to handle this by calling \texttt{git show} and saving the output to a temp file to be used with the AST differencing tool. Here we treated everything as an addition. Since we compared each commit to its parent, as reported by Git, we were able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, this indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{data-extracted}
\caption{Hierarchy of data extracted by TypeV. Each repository may have zero or more commits; each commit may have zero or more changed files; each changed files may have zero or more changes, as given by our AST differencing tool.}
\label{fig:data-extracted}
\end{figure}

The data collected from running the AST differencing component over the commits was saved to a file to be visualized later. In total, the information collected from one repository includes all commits including metadata for each commit such as its author, date, SHA (or revision ID), and commit message. Each commit may be composed one zero or more changed files; for each changed Java file, our tool extracts a change---either a declaration or an invocation---the artefact affected---a type or a method---and whether the change is an addition or a deletion of said artefact. Figure~\ref{fig:data-extracted} summarizes the hierarchy of the data extracted by our tool.

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{Screenshot_antlr4}
\caption{Screenshot showing the main view for TypeV. The data being displayed is from the ANTLR4 repository. Light blue bars \additionsbox{10pt} indicate \textbf{additions}; dark red bars \deletionsbox{10pt} indicate \textbf{deletions}.}
\label{fig:main-view}
\end{figure*}

Although rich in detail, the data is difficult to analyze if displayed in its raw text form; instead, we created an HTML5 web visualization application which is intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}). The source code for TypeV---both the AST differencing component and the web application--- is available online on GitHub.\footnote{\url{https://github.com/mdfeist/TypeV}} A screencast demonstrating its use is also available online.\footnote{\url{https://archive.org/details/typev-demo}}

Since we have the date and authors attached to each commit we are able to group commits together. This allows us to show only changes for a specific author or group of authors. We can also show changes that happened in a specific time period.

TypeV provides a variety of views to help organize the data and provide different levels of detail. The main view displays a series of stacked bar graphs which show changes to types over a period of time in a project (Figure~\ref{fig:main-view}). The light blue  \additionsbox{10pt} and dark red \deletionsbox{10pt} portions account for the number of additions and deletions respectively, in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). The height of each bar shows the number of changes to a type. The width of each bar represents a set time period and can be changed to an hour, day, week, or month. A bar which fills the row vertically is the period with the largest number of changes to that type out of the visible time periods. The other bars in the row are proportional to that maximum number of changes.  The bar height is calculated by multiplying the maximum bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. The rows are sorted so that the graph for the most used types appear the top and are in descending order.

When a user mouses over a specific bar in the graph they are given information for the type and time period at that location. The information includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message.

One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to the Git configuration. This may result in one author appearing as several different users in the \texttt{git log}. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple users at a time, thus allowing the user to view changes from an individual or group of developers.

To allow for a more detailed analysis of a specific component of the repository we added some basic filtering tools. First a user can filter by type name. Second the user can view changes that fall within a specific start and end date. Third the user can specify the time span represented by each bar: hour, day, week, or month. Finally, the user can choose to see only declarations of types, types used in declarations and invocations, or only invocations.

@@

\begin{table}[t]
\renewcommand{\arraystretch}{1.3}
\centering
\caption{Summary of projects used in the evaluation.}
\label{tab:project-summary}
\begin{tabular}{c|lllll}
\hline
\bfseries Project & \bfseries Commits & \bfseries Authors* & \bfseries Files &   \bfseries Types &  \bfseries  Diffs\\
\hline
ANTLR4 & 3930 & 27 & 1793 & 2635 & 58425\\
BookKeeper & 660 & 12 & 758 & 1872 & 31817\\
Curator & 1524 & 43 & 1211 & 1269 & 27957\\
Tika & 860 & 10 & 604 & 757 & 20764\\
\hline
\multicolumn{6}{p{\linewidth}}{\footnotesize Totals for the four projects we analyzed. ``Diffs'' refers to the total number of invocation and declaration changes observed from the AST. Note that the number given for authors may be larger than the actual number of collaborators on the project, due to some collaborators using multiple email address, and thus appearing as more than one author.}
\end{tabular}
\end{table}

\begin{table}[t]
\renewcommand{\arraystretch}{1.3}
\caption{Comparison of Type Coverage and File Coverage in Projects}
\label{tab:summary}
\centering
\begin{tabular}{c|ccc}
\hline
\bfseries Project & \bfseries Wilcoxon& \bfseries Pearson (linear) & \bfseries Spearman (rank) \\
& $p$-value & Correlation & Correlation \\
\hline
ANTLR4 & $p\ll0.01$ & 0.943 & 0.907\\
BookKeeper & $0.313$ & 0.984 & 0.972\\
Curator & $0.007$ & 0.975 & 0.980\\
Tika & $0.653$ & 0.955 & 0.998\\
\hline
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{antlr-density}
\caption{}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{tika-density}
\caption{}
\end{figure}

In Java, a type can be a primitive, provided by the language itself, or an object reference. Primitive types are those that are built into the language such as an \texttt{int} or \texttt{char}. Object references point to instances of classes. Java's syntax encourages the creation of classes, thus we expect substantial Java projects to define many domain-specific types. Additionally, Java places the constraint that each source file must define one class. When writing or contributing to a project, the author's choice of which part of the program to edit will in turn affect what types they end up working with --- that is, it will affect the author's \emph{type coverage}. Similarly, the files an author works with

>>>>

\begin{abstract}
What is a developer's contribution to a repository? By only counting commits and number of lines changed, existing tools that visualize source code repositories (such as GitHub's graphs) fall short on showing the effective contributions made by each developer. When many commits are viewed as a group, the details are lost. Commit information can be misleading since lines of code give no indication of what was actually being worked on without careful examination of the changed code. Providing a semantic view of this information could provide deeper insights into how software projects evolve since changes to design and features are not clearly visible from line changes alone. We present TypeV: a method for visualizing Java source code repositories. Instead of counting line changes in a commit we extract detailed type information over time by using the differences between abstract syntax trees (ASTs). We are then able to track the additions and deletions of declarations and invocations for each type. Furthermore, we can track each author's type usage over time. Using TypeV, we examine specific cases in well-known repositories where our tool reveals interesting and useful information. We then compare type coverage information from the AST compared to file coverage to determine if unique information is provided by type information.
\end{abstract}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{antlr4-code-freq}
\caption{GitHub's code frequency graph for ANTLR4. Note that the graph was cropped to better show the details. Thus, the changes that occur in mid 2015 are much larger than shown. The additions reach over 400k and the deletions are over 300k.}
\label{fig:GH_code_freq}
\end{figure*}

@@

Source code is often represented as text in a source code file. This text is then parsed by a compiler or interpreter with the intent of eventually executing the code represented by the source text. While programmers may type in individual characters, their intention is at a much higher level. Their intent is written in tokens that represent statements, identifiers, and types that are used to solve their problems. Programmers solve problems in a way that is readable to other developers and executable by a computer. The programmer rarely operates at the character level, or even the line level, so why should our analyses of source code operate at granularities unfamiliar to developers?
The plain text nature of software does not stop at source code level. The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which store the differences in lines of text in a file between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for Git repositories provide some basic tools for analyzing projects over time~\cite{github-graphs,bitbucket-graphs}; however, as we will show, these tools often lack the kind of detail to gather insightful analysis. As with source code, commits operate at the text level. While commits may store individual changes as lines of code, they imply a change in \emph{structure} of the software.  By visualizing lines changed, or files changed, the current state-of-the-art for software project visualization rarely capture the \emph{meaning} behind each commit; lines of code alone does not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed does not necessarily reflect meaningful changes to a repository~\cite{robles2014}.

Consider GitHub graphs~\cite{github-graphs}, visualizations that are easily available to open source software developers. GitHub's graphs use line and commit information exclusively. In the case of the ANTRL4 project (Figure~\ref{fig:GH_code_freq}), this code frequency chart creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to crop the graph to be able to show details of other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by a number of reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps a massive refactoring effort \cite{Hindle:2008:LCT:1370750.1370773}.

@@

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualize changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

We will examine research which analyses software repositories for information on language and features usage, tools for visualizing project evolution, collaboration in software repositories and other tools for understanding repositories.

@@

Analyzing source code is a popular research topic and Java is often chosen as the subject of study due to its popularity. Grechanik \textit{et al.} \cite{Grechanik:2010:EIL:1852786.1852801} examined the structure of Java programs mined from 2080 programs. This paper examined the breakdown of syntactic structures in open source repositories, however it does not consider per author statistics. Parnin \textit{et al.} \cite{Parnin:2011:JGA:1985441.1985446} mined repositories to see how Java generics have been used in open source projects and found that generic usages were often introduced by a single developer in a project and were primarily being used for collecting and traversing lists of objects.

Further analysis of the Java language has been done using ASTs. L\"{a}mmel \textit{et al.} \cite{Lammel:2011:LAA:1982185.1982471} used ASTs to examine the usage of APIs in Java projects. They found what APIs are popular and if they were used in a framework-like manner. Dyer \textit{et al.} \cite{Dyer:2014:MBA:2568225.2568295} mined AST nodes to study the use of new Java language features over time. The authors found the most popular features and the adoption rate of new features over time. They did not check to see how much a developer used a feature but instead checked to see if they used it at all. In a more general analysis of programming languages, Meyerovich and Rabkin \cite{Meyerovich:2013:EAP:2509136.2509515} surveyed developers and examined repositories to learn about language adoption and usage. Developers feel that certain language features are more important than others.

The focus of these papers is on how programming languages and features are used by looking at many repositories. Java is a statically-typed language, yet there is no focus specifically on how types are used throughout a project. Type data may provide additional information on how developers use languages but this has yet to be explored.

@@

A variety of visualizations tools have been created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.} \cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.} \cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies \cite{6980224} or files \cite{6650525, 7332421,voinea2006} over time.

Tools also exist which track changes over time in a more complex manner. Kim \textit{et al.} \cite{Kim:2009:DRS:1555001.1555046} built a tool called Logical Structural Diff (LSdiff) which examines code changes and produces logical rules to represent them. Changes which do not match the rules are then pointed out to the developer. Reiss and Tarvo created an IDE which allows users to see the history of files and changes to lines of code from each author as they are editing a project \cite{6650521}. Van Hees and Hage use a voronoi treemap to visualize changes in class and method structure \cite{7332410}. Servant and Jones created CHRONOS, a tool which creates a timeline of changes to a selection of code as well as the ability to query those changes \cite{6650547}.

Some tools examine the evolution of projects to learn other information about them. Alcocer \textit{et al.} \cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files \cite{6980223}. This allowed testing practices to be more closely examined.

None of the aforementioned tools perform comparisons of ASTs or specifically examine the changes to type information for their view of project evolution.

@@

There is also research on how developers use software repositories to collaborate. Wagstrom \textit{et al.} \cite{Patrick:Wagstrom:2012} explored the roles that developers take in a networked, social development environments like GitHub. These role were defined as their level of contribution as well as the types of issues they dealt with. They found that developers will take on multiple roles in a project and will sometimes fulfill the same role across different projects. Lee \textit{et al.} \cite{lee2013} developed a tool for visualizing the branching structure of Git repositories. This was used to analyze concurrent workflows in popular open source repositories. Elsen also created a visualization for branches in Git, with the ability to expand branches to view the directory structure of a project at that point \cite{6650522}.   Shrestha \textit{et al.} made a visualization for the geographic location of a commit and \cite{6650532} and Minelli \textit{et al.} \cite{6980226} looked at the real time changes developers made to a repository.

@@

Some visualization tools are used for managing or making decisions about software rather than study the changes taking place. The service EasyBi \cite{EasyBi} provides additional visualizations for Git repositories for the purpose of business decisions. EasyBi works by taking the git log file and generating charts and graphs from it. It tries to solve the issue of summarizing many commits and has greater controls for granularity and authors but still operates on commits and line differences. The Gitinspector project \cite{Gitinspector} is an open source tool for the statistical analysis of Git repositories. It provides an advanced timeline view of changes per author and cumulative statistics but again relies upon the line differences to summarize the work of each author.

The basic tools easily available to any software developer are the GitHub graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the type of code which was committed. It does not matter what language is used or if errors exist in the commit, GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section \ref{sec:viz}.

An \emph{abstract syntax tree} (AST) is a data structure derived from source code that breaks its syntactic constructs into a tree. Each node in the tree represents a syntactically valid chunk of code. This captures the structure of the code in an abstract and tractable manner. When a programmer makes a non-trivial change in code in a file, that difference will be reflected in the AST. Non-trivial changes are those that affect how a compiler will interpret the source code, and may have effects on the executable code. Trivial changes are those that have no affect on how a compiler will interpret the source text, such as changing insignificant whitespace or comments in source code. Thus, ASTs of consecutive revisions in a code repository can be compared to see what \emph{structures} a programmer is modifying.

@@

\begin{description}
\item [declaration:] 
  Stating that use of a field or variable of a given type. For example declaration of a \texttt{String} type variable might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{String myString;} \\
  \end{adjustwidth}
\item [invocation:] 
  A call site for a method that belongs to a type. For example an invocation of the trim method of a String type might look as follows:
  \\
  \begin{adjustwidth}{2.5em}{0pt}
  \texttt{myString.trim();} \\
  \end{adjustwidth}
\end{description}

@@

This states that two declarations of type \texttt{java.net.URI} were added, and two declarations of type \texttt{java.lang.String} were deleted. Note the fully qualified type names. The source text may not literally have the characters \texttt{java.net.URI}, but Spoon was able to infer from the import statements in the Java source text that references to ``URI'' resolve to the fully-qualified type name \texttt{java.net.URI}. Thus, TypeV is able to keep track of the actual class being used and is not confused when two classes in separate packages have the same name.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by using calling \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure \ref{fig:commits}. Since we compared each commit to its parent, as reported by Git, we were able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure \ref{fig:commits}, it indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{data-extracted}
\caption{Hierarchy of data extracted by TypeV. Each repository may have zero or more commits; each commit may have zero or more changed files; each changed files may have zero or more changes, as computed by our AST differencing tool.}
\label{fig:data-extracted}
\end{figure}

The data collected from running the AST differencing component over the commits is saved to a file to be visualized later; this data is in a general format, and is not bound to any particular visualization. Figure~\ref{fig:data-extracted} summarizes the hierarchy of the data extracted by our tool.

@@

\label{sec:viz}

\begin{figure*}[!ht]
\centering
\includegraphics[width=\textwidth]{Screenshot_antlr4}
\caption{Screenshot showing the main view for TypeV. The data being displayed is from the ANTLR4 repository. Light blue bars \additionsbox{10pt} indicate \textbf{additions}; dark red bars \deletionsbox{10pt} indicate \textbf{deletions}.}
\label{fig:main-view}
\end{figure*}

Although rich in detail, the AST diff data is difficult to analyze if displayed in its raw form; hence, we created an HTML5 web visualization application which is intended for developers, project managers, and researchers to use (Figure~\ref{fig:main-view}). The source code for TypeV---both the AST differencing component and the web application--- is available online on GitHub.\footnote{\url{https://github.com/mdfeist/TypeV}} A screencast demonstrating its use is also available online.\footnote{\url{https://archive.org/details/typev-demo}}

\subsection{Views}

TypeV provides a variety of views to help organize the data and provide different levels of detail. The overview displays several rows of stacked bar graphs which show changes to a type or method over a period of time in a project (Figure~\ref{fig:main-view}). The purpose of displaying stacked bar charts is that they are a familiar visualization, thus are somewhat easy to read, and can succinctly display the relative amounts of additions and deletions per each time period. The light blue \additionsbox{10pt} and dark red \deletionsbox{10pt} portions account for the number of additions and deletions respectively, in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). These colours were chosen as they are high contrast, and are distinct to those red-green colour blindness.

The height of each bar shows the relative number of changes to a type. The width of each bar is fixed, and represents a time period for binning commits that can be configured to an hour, day, week, or month's worth of commits per each bar.  A bar that completely fills its row vertically is the period with the largest number of changes to that type in the visible date range. The height of the other bars in the row are in proportion to the maximum number of changes. The bar height is calculated by multiplying the maximum bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. The rows are sorted so that the graph for the most used types or methods appear the top and are in descending order of use.

One can interact with the overview by hovering and clicking on the bars. When a user mouses over a specific bar in the graph they are given summary information for the type and time period at that location. The information includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message. By adding interaction, the user is able to explore the overview, find the specific information they are interested and drill-down at their leisure. Although our tool extracts a great deal of information, by using interaction, users can gradual reveal more information as they need it.

To allow for a more detailed analysis of a specific component of the repository we added some basic filtering tools. Users can filter by type name; they can view changes that fall within a specific start and end date; users can also choose to see only declarations of types, types used in declarations and invocations, or only invocations. One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different users in the \texttt{git log}. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple users at a time, thus allowing the user to view changes from an individual or group of developers.

Knowing type and author information enables other kinds of visualizations. For example, Figure \ref{fig:chord} is a chord diagram which shows connections between authors and types. The wider the arc around the circumference of the circle, the more often that entity is connected to other entities. For example, ``parrt'' is the top contributor, with changes covering a large number of types. Determining author and type information is made possible by AST differencing.

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{chord}
\caption{Chord diagram showing connections between authors (right and bottom) to types (left).}
\label{fig:chord}
\end{figure*}

@@

\begin{table}[t]
\renewcommand{\arraystretch}{1.3}
\centering
\caption{Summary of projects used in the evaluation.}
\label{tab:project-summary}
\begin{tabular}{c|lllll}
\hline
\bfseries Project & \bfseries Commits & \bfseries Authors* & \bfseries Files &   \bfseries Types &  \bfseries  Changes\\
\hline
ANTLR4 & 3930 & 27 & 1793 & 2635 & 58425\\
BookKeeper & 660 & 12 & 758 & 1872 & 31817\\
Curator & 1524 & 43 & 1211 & 1269 & 27957\\
Tika & 860 & 10 & 604 & 757 & 20764\\
\hline
\multicolumn{6}{p{\linewidth}}{\footnotesize Totals for the four projects we analyzed. ``Changes'' refers to the total number of invocation or declaration changes observed from the AST. Note that the number given for authors may be larger than the actual number of collaborators on the project, due to some collaborators using multiple email address, and thus appearing as more than one author.}
\end{tabular}
\end{table}

\begin{table}[ht]
\renewcommand{\arraystretch}{1.3}
\caption{Comparison of Type Coverage and File Coverage in Projects}
\label{tab:summary}
\centering
\begin{tabular}{c|ccc}
\hline
\bfseries Project & \bfseries Wilcoxon& \bfseries Pearson (linear) & \bfseries Spearman (rank) \\
& $p$-value & Correlation & Correlation \\
\hline
ANTLR4 & $p\ll0.01$ & 0.943 & 0.907\\
BookKeeper & $0.313$ & 0.984 & 0.972\\
Curator & $0.007$ & 0.975 & 0.980\\
Tika & $0.653$ & 0.955 & 0.998\\
\hline
\end{tabular}
\end{table}

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{antlr-density}
\caption{Density plot of type and file coverage for commits. The graph shows what percentage of type and file coverage the majority of commits have. One can see that the majority of commits have a type and file coverage around 0\% to 20\%. We can also see a difference between these measurements on the higher end of the graph. File coverage seems to have a group of commits that had a file coverage between 20\% and 50\% whereas type coverage have a similar looking grouping but between 40\% and 60\%.}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{tika-density}
\caption{Density of type and file coverage for commits. Here we can see that commits did not have much file coverage between 25\% and 75\%. On the other hand there were a lot of commits that have type coverage around 50\%.}
\end{figure}
